\textbf{ Spectrophotometric parallax:\\ Accurate estimates of luminosity and distance\\ for luminous red-giant stars }

### 2018-08-01 paper/sp.tex
making the notation worse

% math macros
\newcommand{\T}{^{\mathsf{T}}}
\DeclareMathOperator*{\argmin}{argmin}
\newcommand{\logg}{\log g}

@@

\item[sparsity] We expect that only a small subset of the full complement of
\apogee\ spectral pixels will be relevant to the prediction of parallax.
That is, we expect that many of the pixels will or ought to get no weight in the
final model that we use to predict luminosities and distances.

@@

The model and log-likelihood function can be expressed heuristically as
\begin{eqnarray}
\varpi_n &=& \exp(\theta\cdot x_n) + \mbox{noise}
\\
\chi^2(\theta) &\equiv& \sum_{n=1}^N \frac{[\varpi_n - \exp(\theta\cdot x_n)]^2}{\sigma_n^2}
\quad ,
\end{eqnarray}
where
$\varpi_n$ is the \gaia\ measurement (adjusted; see below) of the parallax of star $n$,
the model is that the logarithm of the true parallax
can be expressed as a linear combination of the components
of some $D$-dimensional feature vector $x_n$,
$\theta$ is a $D$-vector of linear coefficients,
$\sigma_n$ is the \gaia\ estimate of the uncertainty on the parallax measurement,
and $\chi^2(\theta)$ is (twice) the negative-log-likelihood for the parameters $\theta$
under the assumption of known Gaussian noise and
that there are $N$ independently measured stars $n$.
The feature vector $x_n$ contains photometry in a few bands, and all 8000-ish pixels
in the pseudo-continuum-normalized \apogee\ spectrum, so $D$ is on the order of 8000.

@@

Once the model is optimized, the prediction for a new stellar parallax
$\varpi_m$ for a star $m$ in the validation or test set is
\begin{eqnarray}
\varpi_m &=& \exp(\theta\cdot x_m)
\quad ,
\end{eqnarray}
where $x_m$ is the feature $D$-vector for star $m$.
Thus the model can be trained on a training set of stars and applied to make
predictions for a validation set or a test set.
The only requirements are that every test-set object $m$  must have a full feature
vector $x_m$ just as every training-set object $n$ must have a full feature
vector $x_n$.

One final detail: We perform all of the fitting in a two-fold train-and-test framework,
in which the data are split into two disjoint subsets, A and B.
The model trained on the A data is used to predict the $\varpi_m$ values (and hence
the distances and parallaxes) of the B data,
and the model trained on the B data is used to 
predict the $\varpi_m$ values of the A data.
This ensures that, in the estimate of any individal star's
parallax, none of the \gaia\ data pertaining to that star were used.
This makes the parallax estimates from the spectrophotometric feature vectors
$x_m$ statistically independent of the \gaia\ parallax measurements.
They are not independent globally---\gaia\ data were used to train the model---but
each star's spectrophotometric parallax estimate is independent
of the astrometric estimate from \gaia\ on a star-by-star basis.

@@

In detail,
for each star $n$ in the Training Sets, we take from \gaia\ DR2 we the parallax $\varpi_n$
and its uncertainty $\sigma_n$.
Importantly, to each \gaia\ parallax measurement we add a positive
offset of $0.029$\,mas to adjust for the under-estimation of
parallaxes reported by the \gaia\ Collaboration (HOGG CITE LINDEGREN).
The training sets are shown in HOGG WHAT FIGURES?

For every star $n$ in the full Parent Sample we construct the feature
vector $x_n$ as
\begin{eqnarray}
x_n\T &\equiv& [1, G_n, ({BP}-{RP})_n, J_n, H_n, K_n, W_{1n}, W_{2n}, f_{1n}, f_{2n}, \cdots, f_{Ln}]
\quad ,
\end{eqnarray}
where the 1 permits a linear offset,
the photometry is from \gaia, \zmass, and \wise, respectively,
and
the fluxes in the $L=$ACE \apogee\ spectral pixels for star $n$ are denoted
$f_{1n}$, $f_{2n}$, and so on.
These feature vectors live in a $D$-dimensional space where $D=$HOGG.
For every star $n$ in either of the Training Sets we additionally require---in
addition to these feature vectors---a parallax $\varpi_n$ and uncertainty $\sigma_n$.

>>>>

% math macros
\newcommand{\T}{^{\mathsf{T}}}
\DeclareMathOperator*{\argmin}{argmin}
\newcommand{\logg}{\log g}
\newcommand{\gparallax}{\varpi^{(\mathrm{a})}}
\newcommand{\sparallax}{\varpi^{(\mathrm{sp})}}

@@

\item[sparsity] We expect that only a small subset of the full complement of
\apogee\ spectral pixels will be relevant to the prediction of parallax
(the spectrophotometric parallax).
That is, we expect that many of the pixels will or ought to get no weight in the
final model that we use to predict luminosities and distances.

@@

The model and log-likelihood function can be expressed heuristically as
\begin{eqnarray}
\gparallax_n &=& \exp(\theta\cdot x_n) + \mbox{noise}
\\
\chi^2(\theta) &\equiv& \sum_{n=1}^N \frac{[\gparallax_n - \exp(\theta\cdot x_n)]^2}{\sigma_n^2}
\quad ,
\end{eqnarray}
where
$\gparallax_n$ is the \gaia\ measurement (adjusted; see below) of the parallax of star $n$,
the model is that the logarithm of the true parallax
can be expressed as a linear combination of the components
of some $D$-dimensional feature vector $x_n$,
$\theta$ is a $D$-vector of linear coefficients,
$\sigma_n$ is the \gaia\ estimate of the uncertainty on the parallax measurement,
and $\chi^2(\theta)$ is (twice) the negative-log-likelihood for the parameters $\theta$
under the assumption of known Gaussian noise and
that there are $N$ independently measured stars $n$.
The feature vector $x_n$ contains photometry in a few bands, and all 8000-ish pixels
in the pseudo-continuum-normalized \apogee\ spectrum, so $D$ is on the order of 8000.

@@

Once the model is optimized, the output of the model is a prediction of the parallax,
or really what we will call the \emph{spectrophotometric parallax}.
This spectrophotometric parallax
$\sparallax_m$ for any star $m$ in the validation or test set is
assigned according to
\begin{eqnarray}
\sparallax_m &\leftarrow& \exp(\hat{\theta}\cdot x_m)
\quad ,
\end{eqnarray}
where
$\hat{\theta}$ is the optimal parameter vector according to (HOGG WHAT EQUATION),
and
$x_m$ is the feature $D$-vector for star $m$.
Thus the model can be trained on a training set of stars and applied to make
predictions for a validation set or a test set.
The only requirements are that every test-set object $m$  must have a full feature
vector $x_m$ just as every training-set object $n$ must have a full feature
vector $x_n$.

One final detail: We perform all of the fitting in a two-fold train-and-test framework,
in which the data are split into two disjoint subsets, A and B.
The model trained on the A data is used to predict or produce
the spectrophotometric parallaxes $\sparallax_m$ values (and hence
the distances and parallaxes) of the B data,
and the model trained on the B data is used to 
produce the $\sparallax_m$ values of the A data.
This ensures that, in the estimate of any individal star's
parallax, none of the \gaia\ data pertaining to that star were used.
This makes the parallax estimates from the spectrophotometric feature vectors
$x_m$ statistically independent of the \gaia\ parallax measurements.
They are not independent globally---\gaia\ data were used to train the model---but
each star's spectrophotometric parallax estimate is independent
of the astrometric estimate from \gaia\ on a star-by-star basis.

@@

In detail,
for each star $n$ in the Training Sets, we take from \gaia\ DR2
the parallax $\gparallax_n$ and its uncertainty $\sigma_n$.
Importantly, to each \gaia\ parallax measurement we add a positive
offset of $0.029$\,mas to adjust for the under-estimation of
parallaxes reported by the \gaia\ Collaboration (HOGG CITE LINDEGREN).
The training sets are shown in HOGG WHAT FIGURES?

For every star $n$ in the full Parent Sample we construct the feature
vector $x_n$ as
\begin{eqnarray}
x_n\T &\equiv& [1, G_n, ({BP}-{RP})_n, J_n, H_n, K_n, W_{1n}, W_{2n}, f_{1n}, f_{2n}, \cdots, f_{Ln}]
\quad ,
\end{eqnarray}
where the 1 permits a linear offset,
the photometry is from \gaia, \zmass, and \wise, respectively,
and
the fluxes in the $L=$ACE \apogee\ spectral pixels for star $n$ are denoted
$f_{1n}$, $f_{2n}$, and so on.
These feature vectors live in a $D$-dimensional space where $D=$HOGG.
For every star $n$ in either of the Training Sets we additionally require---in
addition to these feature vectors---a \gaia-measured parallax $\gparallax_n$
and uncertainty $\sigma_n$.
